---
title: "Picasso Triggerfish Reflectance Analysis Demo"
author: "Grok (xAI)"
date: "March 20, 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(signal)  # For derivatives
library(hyperSpec)  # For spectral angle
library(stats)  # For PCA and statistical tests

# Define consistent colors for patches
patch_colors <- c(
  "White" = "gray90",
  "Black" = "gray20",
  "Orange" = "darkorange",
  "Blue" = "dodgerblue"
)
```

# 1.0 Introduction
This demo explores simulated reflectance data for four color patches (White, Black, Orange, Blue) on Picasso triggerfish (*Rhinecanthus aculeatus*) across 400-800 nm, with five fish, each measured three times per patch. We're mimicking analyses from Kolmann et al. (2021) to see if we can distinguish these patches within a fish—like detectives comparing color fingerprints! The data includes noise (measurement error, SD = 2%), inter-fish variation (±10% brightness, ±20 nm shifts), and one faded fish (30% dimmer) to challenge our methods.

## 1.1 Data Simulation
We created fake data mimicking real triggerfish colors:
- **White**: Bright (~70-90%), peaks ~500 nm (blue-green).
- **Black**: Dull (~10%), flat across wavelengths.
- **Orange**: Bright (~90%), peaks ~620 nm (orange-red).
- **Blue**: Bright (~85%), peaks ~470 nm (blue).

These "color recipes" include randomness to reflect natural variation.

## 1.2 Load Data
```{r load-data}
data <- read.csv("Picasso_Triggerfish_Reflectance.csv")
patches <- unique(data$Patch)  # Our four colors
n_ind <- length(unique(data$Individual))  # Five fish
head(data)
```

**Breakdown**:
- **What**: Loads a table with `Individual` (fish ID), `Patch` (color), `Replicate` (measurement number), `Wavelength` (400-800 nm), `Reflectance` (% brightness).
- **Structure**: 9720 rows (5 fish × 4 patches × 3 replicates × 81 wavelengths)—a long list of data points.
- **Illustration**: Picture a spreadsheet—each fish has a block, split by color, then replicate, with 81 wavelength columns (5 nm steps).
- **Why**: This flexible format lets us analyze by fish, patch, or wavelength, building our detective case.

# 2.0 Exploratory Spectral Analysis
Let's visualize the data to get a feel for our patches' "fingerprints."

## 2.1 Figure 2.1-1: Comparative Spectral Signatures
```{r fig-2.1-1}
fig3_data <- data %>%
  group_by(Patch, Wavelength) %>%
  summarise(Mean = mean(Reflectance), SD = sd(Reflectance), .groups = "drop")

ggplot(fig3_data, aes(x = Wavelength, y = Mean, color = Patch)) +
  geom_line() +
  geom_ribbon(aes(ymin = Mean - SD, ymax = Mean + SD, fill = Patch), alpha = 0.2, linetype = 0) +
  scale_color_manual(values = patch_colors) +
  scale_fill_manual(values = patch_colors) +
  labs(title = "Figure 2.1-1: Comparative Spectral Signatures",
       x = "Wavelength (nm)", y = "Reflectance (%)",
       caption = "Mean reflectance per patch across all fish and replicates (n=15 per point). SD bands show within-patch variation (measurement error + fish differences). Between-fish variation is averaged out. Similar to Kolmann et al. (2021) Figure 3, but we use simulated data with broader curves.") +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: `group_by(Patch, Wavelength) %>% summarise(...)` averages 15 measurements (5 fish × 3 replicates) per patch per wavelength, computes SD (324 rows).
- **Step 2**: Plot curves with SD bands—colors match patches.
- **Why**: Shows typical "fingerprints"—baseline for all analyses.
- **Link to Kolmann**: Like Figure 3 (comparative signatures), but their real fish data had sharper peaks; our simulated curves are broader due to simpler models.
- **Key Takeaway**: Distinct curves suggest patches differ; overlap hints they're similar.

## 2.2 Figure 2.2-1: First Derivatives
```{r fig-2.2-1}
fig5_data <- fig3_data %>%
  group_by(Patch) %>%
  arrange(Wavelength) %>%
  mutate(Deriv = c(NA, diff(Mean) / diff(Wavelength)[1])) %>%
  ungroup()

ggplot(fig5_data, aes(x = Wavelength, y = Deriv, color = Patch)) +
  geom_line() +
  scale_color_manual(values = patch_colors) +
  labs(title = "Figure 2.2-1: First Derivative of Spectral Signatures",
       x = "Wavelength (nm)", y = "dR/dλ",
       caption = "First derivative of mean reflectance per patch (n=15 per point). Shows rate of change across wavelengths; within-patch variation averaged. Matches Kolmann et al. (2021) Figure 5, identifying key change regions (e.g., 470 nm for Blue).") +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: `mutate(Deriv = ...)` calculates slopes between consecutive means (5 nm steps), NA for first point (324 rows).
- **Step 2**: Plot slopes—colors match patches.
- **Why**: Highlights "action zones" (e.g., peaks) where patches differ most.
- **Link to Kolmann**: Direct analog to Figure 5—both show derivative peaks (e.g., ~470 nm), though their real data had more detail.
- **Key Takeaway**: Sharp, unique peaks mean distinguishable patches; overlapping slopes suggest similarity.

# 3.0 Analysis Strategies
Now, let's group our methods by how they tackle patch differentiation.

## 3.1 Spectral Angle and ACE Methods
These use spectral angle ($\alpha = \cos^{-1} \left( \frac{\mathbf{x} \cdot \mathbf{S}}{\|\mathbf{x}\| \|\mathbf{S}\|} \right)$) to compare shape similarity, inspired by Kolmann et al. (2021)'s ACE approach.

### 3.1.1 Figure 3.1-1: ACE Spectral Matching
```{r fig-3.1-1}
# Calculate mean spectra if not already done
mean_spectra <- data %>%
  group_by(Individual, Patch, Wavelength) %>%
  summarise(Mean = mean(Reflectance), .groups = "drop") %>%
  pivot_wider(names_from = Wavelength, values_from = Mean)

patch_means <- data %>%
  group_by(Patch, Wavelength) %>%
  summarise(Mean = mean(Reflectance), .groups = "drop") %>%
  pivot_wider(names_from = Wavelength, values_from = Mean)

spectral_angle <- function(x, s) {
  dot <- sum(x * s, na.rm = TRUE)
  norm_x <- sqrt(sum(x^2, na.rm = TRUE))
  norm_s <- sqrt(sum(s^2, na.rm = TRUE))
  acos(dot / (norm_x * norm_s)) * 180 / pi  # Degrees
}

ace_data <- data.frame(
  Individual = rep(unique(data$Individual), each = length(patches)),
  Patch = rep(patches, n_ind),
  Angle = numeric(n_ind * length(patches))
)

for (i in 1:nrow(ace_data)) {
  ind <- ace_data$Individual[i]
  patch <- ace_data$Patch[i]
  x <- as.numeric(mean_spectra[mean_spectra$Individual == ind & mean_spectra$Patch == patch, -(1:2)])
  s <- as.numeric(patch_means[patch_means$Patch == patch, -1])
  ace_data$Angle[i] <- spectral_angle(x, s)
}

ggplot(ace_data, aes(x = Patch, y = Individual, fill = Angle)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red", name = "Spectral Angle (°)") +
  labs(title = "Figure 3.1-1: ACE Spectral Matching",
       x = "Target Patch", y = "Test Individual",
       caption = "Spectral angles between fish means (n=3 replicates) and patch means (n=15). Within-fish variation averaged; between-fish variation (e.g., fading) increases angles. Simplified from Kolmann et al. (2021) Figure 7—uses angle only, no background covariance.") +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: Average replicates (20 rows × 81 columns) and patches (4 rows × 81 columns).
- **Step 2**: Compute $\alpha$—shape similarity across all bands.
- **Step 3**: Heatmap—low angles = good match.
- **Why**: Tests if fish patches align with "typical" shapes using all 81 bands.
- **Link to Kolmann**: Mirrors Figure 7 (ACE matching), but they used full ACE ($D_{ACE}$) with $\Sigma_b$; we simplify to $\alpha$ due to small sample size (60 spectra vs. 81 bands).
- **Key Takeaway**: Low within-patch angles, high between = distinguishable; uniform angles = not.

### 3.1.2 Figure 3.1-2: Within-Fish Patch Pair Angles
```{r fig-3.1-2}
patch_pairs <- combn(patches, 2, paste, collapse = " vs. ")

angle_data <- data.frame()
for (ind in unique(data$Individual)) {
  for (i in 1:(length(patches)-1)) {
    for (j in (i+1):length(patches)) {
      x <- as.numeric(mean_spectra[mean_spectra$Individual == ind & mean_spectra$Patch == patches[i], -(1:2)])
      s <- as.numeric(mean_spectra[mean_spectra$Individual == ind & mean_spectra$Patch == patches[j], -(1:2)])
      angle <- spectral_angle(x, s)
      angle_data <- rbind(angle_data, 
                          data.frame(Individual = ind,
                                     Patch_Pair = paste(patches[i], "vs.", patches[j]),
                                     Angle = angle))
    }
  }
}

# Kruskal-Wallis and pairwise tests
kw_test <- kruskal.test(Angle ~ Patch_Pair, data = angle_data)
pairwise <- pairwise.wilcox.test(angle_data$Angle, angle_data$Patch_Pair, p.adjust.method = "BH")

ggplot(angle_data, aes(x = Patch_Pair, y = Angle)) +
  geom_boxplot(aes(fill = Patch_Pair)) +
  scale_fill_manual(values = colorRampPalette(unname(patch_colors))(length(unique(angle_data$Patch_Pair)))) +
  labs(title = "Figure 3.1-2: Spectral Angles Between Patch Pairs",
       x = "Patch Pair", y = "Spectral Angle (°)",
       caption = paste("Angles between patch pairs within fish (n=5 fish). Kruskal-Wallis p =", round(kw_test$p.value, 3), ". Pairwise p-values test distinctions. Builds on Kolmann et al. (2021)'s spectral angle concept, but focuses within-fish differences.")) +
  theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

cat("Kruskal-Wallis Test:\n")
print(kw_test)
cat("\nPairwise Wilcoxon Tests (BH-adjusted):\n")
print(pairwise)
```

**Breakdown**:
- **Step 1**: Compute $\alpha$ for all patch pairs within each fish (30 rows).
- **Step 2**: Test differences with Kruskal-Wallis (overall) and Wilcoxon (pairwise).
- **Step 3**: Boxplot—high angles = distinct pairs.
- **Why**: Directly tests within-fish patch separation using all 81 bands.
- **Link to Kolmann**: Uses their spectral angle (Equation 1), but not a direct figure match—extends to statistical testing of pairs, unlike their ACE focus (Figure 7).
- **Key Takeaway**: High angles and low p-values (e.g., < 0.05) = distinguishable; low angles, high p-values = not.

## 3.2 Distance-Based Methods
These measure total spectral differences, including amplitude, across all bands.

### 3.2.1 Figure 3.2-1: Euclidean Distance
```{r fig-3.2-1}
euclidean_dist <- function(x, s) {
  sqrt(sum((x - s)^2, na.rm = TRUE))
}

euclid_data <- ace_data
for (i in 1:nrow(euclid_data)) {
  ind <- euclid_data$Individual[i]
  patch <- euclid_data$Patch[i]
  x <- as.numeric(mean_spectra[mean_spectra$Individual == ind & mean_spectra$Patch == patch, -(1:2)])
  s <- as.numeric(patch_means[patch_means$Patch == patch, -1])
  euclid_data$Angle[i] <- euclidean_dist(x, s)
}
colnames(euclid_data)[3] <- "Distance"

ggplot(euclid_data, aes(x = Patch, y = Individual, fill = Distance)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red", name = "Euclidean Distance") +
  labs(title = "Figure 3.2-1: Euclidean Distance Analysis",
       x = "Target Patch", y = "Test Individual",
       caption = "Euclidean distance between fish means (n=3) and patch means (n=15), using all 81 bands. Within-fish variation averaged; between-fish variation (e.g., fading) increases distance. Not in Kolmann et al. (2021)—complements their shape focus.") +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: Compute Euclidean distance across all 81 bands (20 rows).
- **Step 2**: Heatmap—high distances = less typical.
- **Why**: Captures total difference (shape + amplitude), unlike spectral angle.
- **Link to Kolmann**: Not used in their paper—they focused on shape (ACE); this adds amplitude sensitivity.
- **Key Takeaway**: High between-patch distances = distinguishable; overlap = not.

## 3.3 Principal Component Analysis (PCA)
PCA reduces all 81 bands into key patterns.

### 3.3.1 Figure 3.3-1: PCA of Spectral Signatures
```{r fig-3.3-1}
spectra_matrix <- mean_spectra[, -(1:2)] %>% as.matrix()
pca_result <- prcomp(spectra_matrix, scale. = TRUE)
var_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2) * 100
cumvar_explained <- cumsum(var_explained)

pca_data <- as.data.frame(pca_result$x) %>%
  mutate(Individual = mean_spectra$Individual, Patch = mean_spectra$Patch)

ggplot(pca_data, aes(x = PC1, y = PC2, color = Patch, shape = Individual)) +
  geom_point(size = 3) +
  scale_color_manual(values = patch_colors) +
  labs(title = "Figure 3.3-1: PCA of Spectral Signatures",
       x = paste0("PC1 (", round(var_explained[1], 1), "% variance)"),
       y = paste0("PC2 (", round(var_explained[2], 1), "% variance)"),
       caption = "PCA scores from all 81 bands (n=20 spectra). Within-fish variation averaged; between-fish and within-fish patch differences drive separation. Not directly in Kolmann et al. (2021)—complements their spectral focus.") +
  theme_minimal()

cat("\nVariance Explained by PCs:\n")
print(data.frame(PC = 1:5, Variance_Explained = round(var_explained[1:5], 1), Cumulative = round(cumvar_explained[1:5], 1)))
```

**Breakdown**:
- **Step 1**: `prcomp(...)` reduces 81 bands to PCs (20 rows × 2).
- **Step 2**: Scatter plot—PC1 vs. PC2.
- **Why**: Visualizes clustering using all spectral data.
- **Link to Kolmann**: Not a direct figure match—they used spectral angle/ACE; PCA is a common hyperspectral tool, offering a different lens.
- **Key Takeaway**: Separate clusters = distinguishable; overlap = not.

### 3.3.2 Figure 3.3-2: PCA Loadings
```{r fig-3.3-2}
wavelengths <- as.numeric(colnames(spectra_matrix))
loadings_data <- data.frame(
  Wavelength = wavelengths,
  PC1 = pca_result$rotation[,1],
  PC2 = pca_result$rotation[,2]
) %>%
  pivot_longer(cols = c(PC1, PC2), names_to = "Component", values_to = "Loading")

ggplot(loadings_data, aes(x = Wavelength, y = Loading, color = Component)) +
  geom_line() +
  labs(title = "Figure 3.3-2: PCA Loadings",
       x = "Wavelength (nm)",
       y = "Loading",
       caption = "Loadings show how each wavelength contributes to PC1 and PC2. Peaks highlight diagnostic bands.") +
  theme_minimal()

# Print top contributing wavelengths
top_loadings <- function(rotation, n = 5) {
  abs_loadings <- abs(rotation)
  top_indices <- order(abs_loadings, decreasing = TRUE)[1:n]
  wavelengths <- as.numeric(rownames(rotation))
  data.frame(
    Wavelength = wavelengths[top_indices],
    Loading = rotation[top_indices]
  )
}

cat("\nTop Contributing Wavelengths:\n")
cat("\nPC1:\n")
print(top_loadings(pca_result$rotation[,1]))
cat("\nPC2:\n")
print(top_loadings(pca_result$rotation[,2]))
```

**Breakdown**:
- **Step 1**: Extract loadings—how each wavelength contributes to PCs.
- **Step 2**: Plot loadings and list top contributors.
- **Why**: Identifies which wavelengths drive patch separation.
- **Link to Kolmann**: Not used—they focused on spectral angles; this adds insight into key wavelengths.
- **Key Takeaway**: Strong loadings = diagnostic bands; weak = less important.

### 3.3.3 Figure 3.3-3: Mahalanobis Distance
```{r fig-3.3-3}
mahalanobis_dist <- function(x, mean_vec, cov_mat, lambda = 0.1) {
  n <- nrow(cov_mat)
  regularized_cov <- cov_mat + lambda * diag(n)
  diff <- x - mean_vec
  sqrt(t(diff) %*% solve(regularized_cov) %*% diff)
}

mahal_data <- ace_data
cov_mat <- cov(spectra_matrix)
for (i in 1:nrow(mahal_data)) {
  ind <- mahal_data$Individual[i]
  patch <- mahal_data$Patch[i]
  x <- as.numeric(mean_spectra[mean_spectra$Individual == ind & mean_spectra$Patch == patch, -(1:2)])
  s <- as.numeric(patch_means[patch_means$Patch == patch, -1])
  mahal_data$Angle[i] <- mahalanobis_dist(x, s, cov_mat)
}
colnames(mahal_data)[3] <- "Distance"

ggplot(mahal_data, aes(x = Patch, y = Individual, fill = Distance)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red", name = "Mahalanobis Distance") +
  labs(title = "Figure 3.3-3: Mahalanobis Distance Analysis",
       x = "Target Patch", y = "Test Individual",
       caption = "Mahalanobis distance between fish means (n=3) and patch means (n=15), using all 81 bands with covariance adjustment. Within-fish variation averaged; between-fish variation affects distances. Not in Kolmann et al. (2021)—extends distance with covariance.") +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: Compute Mahalanobis distance with covariance (20 rows).
- **Step 2**: Heatmap—adjusted distances.
- **Why**: Like Euclidean but accounts for band correlations across all 81 bands.
- **Link to Kolmann**: Not used—they used ACE's covariance; this is a related but distinct covariance approach.
- **Key Takeaway**: High between, low within = distinguishable; similar distances = not.

## 3.4 Diagnostic Spectra Methods
These focus on key "diagnostic" bands where patches differ most.

### 3.4.1 Figure 3.4-1: Derivative Peak Analysis
```{r fig-3.4-1}
diag_bands <- fig5_data %>%
  group_by(Patch) %>%
  mutate(Abs_Deriv = abs(Deriv)) %>%
  arrange(desc(Abs_Deriv)) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  distinct(Wavelength, .keep_all = TRUE)

ggplot() +
  geom_line(data = fig3_data, aes(x = Wavelength, y = Mean, color = Patch)) +
  geom_point(data = diag_bands, aes(x = Wavelength, y = Mean, color = Patch), size = 3) +
  scale_color_manual(values = patch_colors) +
  labs(title = "Figure 3.4-1A: Diagnostic Bands from Derivatives",
       x = "Wavelength (nm)", y = "Reflectance (%)",
       caption = "Dots mark diagnostic bands with highest change rates (n=15 per mean). Inspired by Kolmann et al. (2021) Figure 5—targets key wavelengths.") +
  theme_minimal()

diag_spectra <- mean_spectra[, c("Individual", "Patch", as.character(diag_bands$Wavelength))]
diag_means <- patch_means[, c("Patch", as.character(diag_bands$Wavelength))]

deriv_data <- data.frame(
  Individual = rep(unique(data$Individual), each = length(patches)),
  Patch = rep(patches, n_ind),
  Angle = numeric(n_ind * length(patches))
)

for (i in 1:nrow(deriv_data)) {
  ind <- deriv_data$Individual[i]
  patch <- deriv_data$Patch[i]
  x <- as.numeric(diag_spectra[diag_spectra$Individual == ind & diag_spectra$Patch == patch, -(1:2)])
  s <- as.numeric(diag_means[diag_means$Patch == patch, -1])
  deriv_data$Angle[i] <- spectral_angle(x, s)
}

ggplot(deriv_data, aes(x = Patch, y = Individual, fill = Angle)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red", name = "Spectral Angle (°)") +
  labs(title = "Figure 3.4-1B: Derivative Peak Analysis",
       x = "Target Patch", y = "Test Individual",
       caption = paste("Spectral angles using", nrow(diag_bands), "diagnostic bands (e.g.,", paste(head(diag_bands$Wavelength, 3), collapse = ", "), "...). Within-fish variation averaged; between-fish variation increases angles.")) +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: Identify 5 peak derivative bands per patch, deduplicate (~15 bands).
- **Step 2**: Plot bands on spectra (A); compute spectral angle on subset (B).
- **Why**: Focuses on "action zones" (e.g., 470 nm) instead of all 81 bands.
- **Link to Kolmann**: Builds on Figure 5 (derivatives)—they used this to find key regions, we extend to angle analysis.
- **Key Takeaway**: Low within, high between = distinguishable with key bands; similar = not.

### 3.4.2 Figure 3.4-2: Variance-Based Band Analysis
```{r fig-3.4-2}
var_data <- mean_spectra %>%
  pivot_longer(cols = -c(Individual, Patch), names_to = "Wavelength", values_to = "Reflectance") %>%
  group_by(Wavelength) %>%
  summarise(Variance = var(Reflectance), .groups = "drop") %>%
  arrange(desc(Variance)) %>%
  slice_head(n = 10)

var_data_plot <- fig3_data %>% dplyr::filter(Wavelength %in% var_data$Wavelength)

ggplot() +
  geom_line(data = fig3_data, aes(x = Wavelength, y = Mean, color = Patch)) +
  geom_point(data = var_data_plot, aes(x = Wavelength, y = Mean, color = Patch), size = 3) +
  scale_color_manual(values = patch_colors) +
  labs(title = "Figure 3.4-2A: High-Variance Diagnostic Bands",
       x = "Wavelength (nm)", y = "Reflectance (%)",
       caption = "Dots mark 10 bands with highest variance across patches (n=20 spectra). Inspired by Kolmann et al. (2021)'s variance focus (e.g., Figure S3).") +
  theme_minimal()

var_spectra <- mean_spectra[, c("Individual", "Patch", var_data$Wavelength)]
var_means <- patch_means[, c("Patch", var_data$Wavelength)]

var_angle_data <- data.frame(
  Individual = rep(unique(data$Individual), each = length(patches)),
  Patch = rep(patches, n_ind),
  Angle = numeric(n_ind * length(patches))
)

for (i in 1:nrow(var_angle_data)) {
  ind <- var_angle_data$Individual[i]
  patch <- var_angle_data$Patch[i]
  x <- as.numeric(var_spectra[var_spectra$Individual == ind & var_spectra$Patch == patch, -(1:2)])
  s <- as.numeric(var_means[var_means$Patch == patch, -1])
  var_angle_data$Angle[i] <- spectral_angle(x, s)
}

ggplot(var_angle_data, aes(x = Patch, y = Individual, fill = Angle)) +
  geom_tile() +
  scale_fill_gradient(low = "blue", high = "red", name = "Spectral Angle (°)") +
  labs(title = "Figure 3.4-2B: Variance-Based Band Analysis",
       x = "Target Patch", y = "Test Individual",
       caption = paste("Spectral angles using 10 high-variance bands (e.g.,", paste(head(var_data$Wavelength, 3), collapse = ", "), "...). Within-fish variation averaged; between-fish variation increases angles.")) +
  theme_minimal()
```

**Breakdown**:
- **Step 1**: Select 10 high-variance bands across all spectra.
- **Step 2**: Plot bands on spectra (A); compute spectral angle on subset (B).
- **Why**: Targets natural variation hotspots instead of all 81 bands.
- **Link to Kolmann**: Echoes their variance analysis (Figure S3) for band selection, extended here to angle-based comparison.
- **Key Takeaway**: Low within, high between = distinguishable with key bands; similar = not.

# 4.0 Notes for Undergrads
- **Visual Aids**: Sketch:
  - 2.1-1: Four curves with bands.
  - 2.2-1: Slopes on curves.
  - 3.1-x: Heatmaps (angles) or boxplots (pairs).
  - 3.2-x: Heatmaps (distances).
  - 3.3-1: Dots in clusters.
  - 3.4-x: Curves with diagnostic dots, heatmaps.
- **Key Takeaway**: We're hunting patch "fingerprints" within fish. Big differences (angles, distances, clusters) and low p-values = distinguishable. Small differences = too alike. Each method offers a lens—spectral angle for shape, distance for total difference, PCA for patterns, diagnostic bands for key clues.

# 2.0 Data Simulation ----
# First, simulate Picasso triggerfish data
source("simulate_picasso_data.R")

# Next, simulate Hawaiian fish data
source("simulate_hawaiian_fish.R")
# This creates a dataset with reflectance measurements for 51 Hawaiian reef fish species
# The data is saved to "Hawaiian_Reef_Fish_Reflectance.csv"

# 5.0 Hawaiian Fish Spectral Analysis
## 5.1 Load and Process Hawaiian Fish Data
```{r load-hawaiian-data}
hawaiian_data <- read.csv("Hawaiian_Reef_Fish_Reflectance.csv")

# Define color mapping including UV representation
hawaiian_colors <- c(
  "UV" = "darkviolet",
  "Violet" = "mediumorchid",
  "Blue" = "dodgerblue",
  "Green" = "seagreen",
  "Yellow" = "gold",
  "Orange" = "darkorange",
  "Red" = "firebrick",
  "Brown/Olive" = "olivedrab",
  "Black" = "gray20",
  "White" = "gray90",
  "Blue/UV" = "dodgerblue4",
  "Yellow/UV" = "darkgoldenrod",
  "Orange-UV" = "darkorange4",
  "Red/UV" = "darkred",
  "Blue/red" = "mediumvioletred",
  "Labriform-green" = "mediumseagreen",
  "Labriform-purple" = "mediumpurple"
)
```

## 5.2 Figure 5.2-1: Hawaiian Fish Mean Spectra
```{r fig-5.2-1, fig.height=130, fig.width=20, out.width='100%'}
# Calculate mean spectra per species and patch
hawaiian_means <- hawaiian_data %>%
  group_by(Species, Patch, Wavelength) %>%
  summarise(
    Mean = mean(Reflectance),
    SD = sd(Reflectance),
    .groups = "drop"
  )

# Create the plot
ggplot(hawaiian_means, aes(x = Wavelength, y = Mean, color = Patch, fill = Patch)) +
  geom_ribbon(aes(ymin = Mean - SD, ymax = Mean + SD), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  scale_color_manual(values = hawaiian_colors) +
  scale_fill_manual(values = hawaiian_colors) +
  facet_wrap(~ Species, ncol = 2) +
  labs(
    title = "Hawaiian Fish Spectral Signatures",
    x = "Wavelength (nm)",
    y = "Reflectance (%)",
    caption = "Mean reflectance per patch across individuals (n=3 per point). SD bands show within-species variation.\nUV patches shown in deep purple, complex patches (e.g., Blue/UV) shown in darker variants."
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.caption = element_text(size = 12),
    legend.position = "top",
    panel.grid.major = element_line(color = "gray90", linetype = "dotted"),
    panel.grid.minor = element_line(color = "gray95", linetype = "dotted")
  )
```

**Breakdown**:
- **Step 1**: Load Hawaiian fish data and define color mapping for all patch types.
- **Step 2**: Calculate mean spectra per species and patch (averaging across individuals and replicates).
- **Step 3**: Create multi-panel plot (2 columns) showing each species' patches.
- **Why**: Provides overview of spectral diversity across Hawaiian fish species.
- **Key Features**:
  - UV patches shown in deep purple
  - Complex patches (e.g., Blue/UV) shown in darker variant of primary color
  - Grid lines aid in reading values
  - Legend at top shows patch types
  - Species names as panel titles
  - Standard deviation bands show variation
- **Note for Students**: Notice how different species use different combinations of patches, and how some patches (like UV) occur in specific wavelength regions.

# 6.0 Claude's Simulated Fish Analysis
## 6.1 Load and Process Simulated Fish Data
```{r load-claude-data}
claude_data <- read.csv("Simulated_Fish_Reflectance.csv")

# Define color mapping for simulated patches
claude_colors <- c(
  "UV" = "darkviolet",
  "Violet" = "mediumorchid",
  "Blue" = "dodgerblue",
  "Green" = "seagreen",
  "Yellow" = "gold",
  "Orange" = "darkorange",
  "Red" = "firebrick",
  "Brown/Olive" = "olivedrab",
  "Black" = "gray20",
  "White" = "gray90"
)
```

## 6.2 Figure 6.2-1: Simulated Fish Mean Spectra
```{r fig-6.2-1, fig.height=130, fig.width=20, out.width='100%'}
# Calculate mean spectra per species and patch
claude_means <- claude_data %>%
  group_by(Species, Patch, Wavelength) %>%
  summarise(
    Mean = mean(Reflectance),
    SD = sd(Reflectance),
    .groups = "drop"
  )

# Create the plot
ggplot(claude_means, aes(x = Wavelength, y = Mean, color = Patch, fill = Patch)) +
  geom_ribbon(aes(ymin = Mean - SD, ymax = Mean + SD), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  scale_color_manual(values = claude_colors) +
  scale_fill_manual(values = claude_colors) +
  facet_wrap(~ Species, ncol = 2) +
  labs(
    title = "Claude's Simulated Fish Spectral Signatures",
    x = "Wavelength (nm)",
    y = "Reflectance (%)",
    caption = "Mean reflectance per patch across individuals (n=3 per point). SD bands show within-species variation.\nUV patches shown in deep purple."
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 10),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 16, hjust = 0.5),
    plot.caption = element_text(size = 12),
    legend.position = "top",
    panel.grid.major = element_line(color = "gray90", linetype = "dotted"),
    panel.grid.minor = element_line(color = "gray95", linetype = "dotted")
  )
```

## 6.3 Figure 6.3-1: Spectral Diversity Analysis
```{r fig-6.3-1}
# Calculate spectral diversity metrics
spectral_diversity <- claude_data %>%
  group_by(Species) %>%
  summarise(
    Mean_Reflectance = mean(Reflectance),
    SD_Reflectance = sd(Reflectance),
    Max_Reflectance = max(Reflectance),
    Min_Reflectance = min(Reflectance),
    Num_Patches = n_distinct(Patch),
    .groups = "drop"
  )

# Create a summary plot
ggplot(spectral_diversity, aes(x = reorder(Species, Mean_Reflectance))) +
  geom_point(aes(y = Mean_Reflectance, color = "Mean"), size = 3) +
  geom_errorbar(aes(ymin = Mean_Reflectance - SD_Reflectance, 
                    ymax = Mean_Reflectance + SD_Reflectance,
                    color = "SD"), width = 0.2) +
  geom_point(aes(y = Max_Reflectance, color = "Max"), size = 2) +
  geom_point(aes(y = Min_Reflectance, color = "Min"), size = 2) +
  scale_color_manual(values = c("Mean" = "blue", "SD" = "gray", 
                               "Max" = "red", "Min" = "green")) +
  labs(
    title = "Spectral Diversity Across Species",
    x = "Species",
    y = "Reflectance (%)",
    color = "Metric"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )
```

## 6.4 Figure 6.4-1: Patch Distribution Analysis
```{r fig-6.4-1}
# Calculate patch distribution
patch_distribution <- claude_data %>%
  group_by(Species, Patch) %>%
  summarise(
    Mean_Reflectance = mean(Reflectance),
    .groups = "drop"
  )

# Create a heatmap of patch reflectance
ggplot(patch_distribution, aes(x = Patch, y = Species, fill = Mean_Reflectance)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(
    title = "Patch Reflectance Distribution",
    x = "Patch Type",
    y = "Species",
    fill = "Mean Reflectance (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )
```

**Breakdown of Claude's Simulation Analysis**:
- **Data Structure**: The simulation generated data for 51 species with 3 individuals each, covering 10 different patch types across wavelengths from 400 to 800 nm.
- **Visualization Approach**: 
  - Figure 6.2-1 shows detailed spectral signatures for each species and patch type
  - Figure 6.3-1 provides an overview of spectral diversity across species
  - Figure 6.4-1 offers a heatmap view of patch distribution
- **Key Features**:
  - Consistent color scheme with the Hawaiian fish analysis
  - Standard deviation bands to show variation
  - Faceted layout for easy species comparison
  - Multiple visualization approaches for different aspects of the data
- **Note for Students**: Compare these simulated spectra with the real Hawaiian fish data to understand how well the simulation captures natural variation in fish coloration.